Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-03 analysis/paper/paper.Rmd
Introducing oxcal and more proper treatment of dates

```{r libraries, eval = TRUE}
library(tidyverse)
library(GGally)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(sf)
library(rnaturalearth)
```

```{r get-functions, include = FALSE}
source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
```

```{r get-data, eval = TRUE, results = 'hide'}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))

@@

gis <- read.csv(here::here('../../site_data.csv'))
gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
```

<!-- Here's some example analysis code: -->

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean age for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100)
```

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is also partially informed by the so-called Frison effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

@@

While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies @barton1999 and collegues [e.g. @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context. This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is a higher number of cores and unretouched flakes and blades [@clark2017].

@@

```{r lithics}
lithics <- sites %>%
  # Combine flint projectiles into a single variable
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
    small_tool_other = knife_other + scraper_other + drill_other + 
      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
      flake_retouch_other,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "site") %>% 
  # Deselect a few of the columns not be used
  select(-one_of(c("n", "excavation_project", "excavated_m3")))
```

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370].

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other) %>% 
  column_to_rownames("site")

lithics3 <- gisjoin %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3,
         viewsize_5k = log10(viewsize_5k),
         viewsize_10k = log10(viewsize_10k),
         fetch = log10(avg_fetch)) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other, -cat, -municipality, -quality,
         -elev_minimum, -site_size, -cent_elev, -phase, -slope_average, -aspect_average,
         -infiltration_mode,-emergence_50, -emergence_500, -emergence_1k, -location,
         -island_size) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip) %>% 
  column_to_rownames("site")

@@

@@

```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
# Scree plot indicates an 'elbow' at the second dimension. 
# The two first dimension account for 68 % and the first three
# for 87 % (see table below)
# fviz_screeplot(pcalithics, addlabels=TRUE)

# Create a table showing the eigenvalues of the components
round(as.tibble(get_eig(pcalithics)), 2) %>% 
  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
  rownames_to_column(var = "Component") %>% 
gt(rowname_col = "row", auto_align = FALSE)
```

@@

# Discussion

The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. This is evident in the present data as well, highlighted by the influence of the later Mesolithic axe production sites, which it would be reasonable to assume corresponds to a more pervasive cultural change than a purely typo-chronological development of artefact morphology. Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).

>>>>

```{r libraries, eval = TRUE}
library(tidyverse)
library(GGally)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

```{r get-functions, include = FALSE}
source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())

#setOxcalExecutablePath("/tmp/RtmpFVOcBJ")
```

```{r get-data, eval = TRUE, results = 'hide'}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
radiocarbon <- read.csv(here::here('analysis/data/raw_data/radiocarbon_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))

@@

# gis <- read.csv(here::here('../../site_data.csv'))
# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))
# gisjoin <- right_join(gis, sites, by = c("askeladden_id" = "askeladden_id"))
```

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100)

h3 <- radiocarbon %>% 
  filter(site == "Hovland 3")

h3dat <- R_Date(h3$lab_code, h3$c14_bp, h3$error)
h3_sum <- oxcal_Sum(h3dat)
h3_run <- executeOxcalScript(h3_sum)
h3_text <- readOxcalOutput(h3_run)
h3_result <- parseFullOxcalOutput(h3_text)

oxcal_code <- paste(' Plot()
   {
    Sequence()
    {
     Boundary("Start");
     Phase()
     {',
      h3dat,
     '};
     Boundary("End");
    };
  };')
my_result_file <- executeOxcalScript(oxcal_code)
my_result_text <- readOxcalOutput(my_result_file)
my_result_data <- parseOxcalOutput(my_result_text)

my_oxcal_code <- ' Plot()
 {
  Sequence("Sequence1")
  {
   Boundary("Beginn");
   Phase("Phase1")
   {
    R_Date("Lab-1",5000,25);
    R_Date("Lab-2",4900,37);
   };
   Boundary("Between");
   Phase("Phase2")
   {
    R_Date("Lab-3",4800,43);
   };
   Boundary("End");
  };
 };'

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [@glørstad2010]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

The first part of the analysis conducted here involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods when the entirety of the assemblages are taken into account. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

@@

While the definition of curation, its archaeological correlates, and behavioural implications have been widely discussed [e.g. @bamforth1986, @shott1996], following not least from the ambiguous definition first put forward by @binford1979, some concrete operationalisation of the terms have been suggested. Through a series of studies Barton and collegues [e.g. -@barton1999; @barton2014; @clark2017] have shown that the relationship between lithic volumetric density and relative frequency of retouched artefacts provide consistent results across a wide range of chronological and cultural context. This measure is taken to reflect degree of curation associated with the assemblages, which in turn is mainly to follow from variation in mobility patterns associated with the sites [@barton2014]. Higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliabilty, repairability, and the degree to which artefacts could be manipulated to fulfill a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is in this model assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics and a low relative frequency of retouched artefacts. A further empirical expectation of an expendient tehcnological organisation is a higher number of cores and unretouched flakes and blades [@clark2017].

@@

```{r lithics}
lithics <- sites %>%
  # Combine flint projectiles into a single variable
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
    small_tool_other = knife_other + scraper_other + drill_other + 
      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
      flake_retouch_other,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "site") %>% 
  # Deselect a few of the columns not be used
  select(-n, -excavation_project, -excavated_m3, -dating_method)
```

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. These findings are in large part in line with previous research [e.g. @bang-andersen2003], although possibly highlighting the importance of the non-flint materials to a larger extent than what has traditionally been the case.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. While the outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370, @glørstad2011].

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other, -dating_method) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip, dating_method) %>% 
  column_to_rownames("site")

@@

To visualise this more clearly, a curation index based on the two variables was devised by perfoming a min-max normalisation on the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. Plotting the results indicates a higher degree of curation among the earliest sites, which then falls and settles without any further major fluctuations.

@@

```{r eigenvalues, fig.cap = "Table showing the eigenvalues of the components."}
# Create a table showing the eigenvalues of the components
round(as.tibble(get_eig(pcalithics)), 2) %>% 
  rename("Eigenvalue" = eigenvalue, "Variance (%)" = variance.percent, 
         "Cumulative variance (%)" = cumulative.variance.percent) %>% 
  rownames_to_column(var = "Component") %>% 
gt(rowname_col = "row", auto_align = FALSE)
```

@@

```{r index}
# Create curation index
minmax <- lithics2 %>% 
  mutate(dens =  ((lithic_dens - min(lithic_dens)) / (max(lithic_dens) - min(lithic_dens))),
         secondary = ((secondary_freq - min(secondary_freq)) / (max(secondary_freq) - min(secondary_freq))))

indexdata <- minmax %>% 
  rowwise() %>% 
  mutate(index = mean(c(dens * -1, secondary)))

w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BP",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic() 
w3
```
# Discussion

The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites. Unpicking and aligning the signficance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we lack even a most basic understanding of what any individual lithic object has been used for, leaving the behavioural significance of the employed units of analysis unclear (see section, above).
